Accelerating Burn Speed of Birch Wood Popsicle Sticks

### 2022-05-01 paper/paper.Rmd
Add paper edits and change Skillings Mack to use Bonferroni adjustments

---
title: "Accelerating Burn Speed of Birch Wood Popsicle Sticks"
author: "E. Croffoot-Suede, S. Kim"
date: "4/25/2022"
output:
  pdf_document:
    pandoc_args: --listings
    includes:
      before_body: title.sty
      in_header: preamble.tex
---

@@

Curiously, all three solutions had warnings on the bottles warning of the flammability of each liquid, which bode well for our hypothesis that soaking objects in them would increase that object's flammability.

### Hypotheses

We predicted that using any liquid accelerant would be faster than no liquid accelerant, and that in comparing individual treatments, there would be no significant difference between the accelerants, but there would be a significant difference between no accelerant and with accelerant.

## Methods

@@

For the experiment, we used birch wood popsicle sticks. We also needed our treatments, which were the ethanol alcohol, isopropyl alcohol, acetone, and no accelerant. To hold the burning sticks off the ground (so they could be lit from below - flames rise), we used four mason jars, one for each stick within a block. To light the sticks, we used a barbeque lighter, and for timing, we used an Apple iPhone. For safety purposes, we also had extra water.

@@

### Steps

To ensure that we could find statistically significant results, we ran a power calculation. We assumed that we would use six time blocks, and we chose an alpha level for this of $\alpha = 0.05$, which is the probability of a Type I error (false rejection). We chose a Type II error probability of 0.1. Given a $k = 4$ (given four treatment), the calculated sample size was 20, but given a $k = 6$ (given six time blocks), our calculated sample size was 24. Given our two-way layout, we should take the larger $N$ calculated out of $k = 4, 6$. Based on this, we used $N = 24$, meaning that we should use 6 blocks of 4 observations each.

@@

This method of block generation, however, means that it is possible to have multiple iterations of one treatment and none of other treatments in the same block. For example, it is possible to have "acetone", "acetone", "isopropyl", and "dry" in a single block, meaning that "ethanol" would not be tested in this block. This gives us arbitrary unbalanced data.

### Statistical Analysis

Data were cleaned using Python version 3.7 (Python, Wilmington, DE) and pandas version 1.0.3. Data were analyzed using R version 4.0.3 (The R Project for Statistical Computing, Indianapolis, IN). The Skillings-Mack procedure was run in R using the NSM3 package version 1.16. The Wilcoxon rank-sum tests were run using the stats package in R.

## Analysis

@@

Figure 1 shows the range of response values, but to get a better view of the treatment effects, we also made Figure 2, a plot of unburnt distance by accelerant. We are not using the blocking factor as a factor of interest, so the only plot to be made using a factor against response is with treatment, as shown in Figure 2.

@@

Because of its higher relative asymptotic efficiency, we began by running an analysis of variance on the data. We believe that the independence condition is met because each popsicle stick should not affect another popsicle stick during the burn. However, when we checked the conditions for an analysis of variance, we found that equal variance condition and the normality condition were not met for a parametric test.

```{R normality, fig.show="hold", out.width="50%", fig.height = 5, echo = FALSE}
plot(aov(distance ~ type * run, data = fire), which = 1, main = "Figure 3: Residuals vs Fitted")
plot(aov(distance ~ type * run, data = fire), which = 2, main = "Figure 4: Normal Q-Q")
```

@@

In Figure 4, there are flaired tails that indicate an overdispersed data distribution compared to a normal distribution. That is the distribution was "squished" toward the center compared to the normal distribution.

@@

Thusly, our model is updated to

@@

The Skillings-Mack procedure requires that the $N$ random variables are mutually independent. That is, $X_{ij1}, ..., X_{ijc_{ij}}$, $i = 1, ..., n$, and $j = 1, ..., k$ are mutually independent. We believe this to be true for our data given the collection method. It also requires that for each fixed $(i, j)$, the $c_{ij}$ random variables are a random sample from a continuous distribution function $F_{ij}$. We also believe this to be true because we used a random sample out of the population of popsicle sticks in the world. Lastly, the treatment response should be a continuous variable such that "less than", "equal to", and "greater than" can be placed into context. We believe that "length" in millimeters as a response variable satisfies this condition.

@@

When we ran the Skillings-Mack procedure, we found an asymptotic upper-tail probability of $0.0536 < 0.1 = \alpha$, and a Skillings-Mack SM test statistic of 7.6587 on 3 degrees of freedom. Given this, we concluded that there is a difference in at least one of the groups.

In order to find which groups have a significant difference, we need to use an associated pairwise comparison procedure that can also be used against arbitrary incomplete block designs. To determine this, we created a new pairwise multiple comparisons procedure that can be used on arbitrary incomplete block designs.

1. Run a Wilcoxon rank-sum test on the data for all $6 = {4\choose2}$ treatment pairwise comparisons.
2. Keep the block and response values together, but randomly assign a new treatment for each observation, then calculate the Wilcoxon rank sum for all $6 = {4\choose2}$ treatment pairwise comparisons. This will spread the block effect across all treatments. Repeat this $r$ times, and store each of the pairwise comparisons for each of these tests.
3. For each of the six pairwise comparisons groups, compare the \nth{5} and \nth{95} percentiles of the $r$ simulated random-treatment Wilcoxon rank-sum test statistics against the test statistic from each of the pairwise Wilcoxon tests from the actual data. If the actual Wilcoxon rank-sum test statistic falls inside of the (0.05, 0.95) percentile from the random-treatment distribution, there is a significant difference between the two groups.

This procedure is used to determine that given two groups (in this case, each pair of treatments), is there strong statistical evidence to support the hypothesis that one of the groups is significantly different from the other? Because we are using a nonparametric method, we are comparing the medians. This diminishes the effect of using non-normal data. The Wilcoxon rank-sum test calculates distance from the medians using the sum of the rank, not the actual value's difference from the median, which also helps to diminish the effect of extraordinarily large or small outliers.

@@

We see that acetone has the lowest box by distance, so we would like to take acetone or anything in acetone's group as calculated by the multiple comparisons procedure.

@@

However, we believe that our pairwise multiple comparisons procedure would still be valid given data obtained in an arbitrary incomplete block design. For example, suppose we were looking at data for a factory that produces products in batches, but which does not produce every batch every day. This would generate data in an arbitrary incomplete block design, where a metric of the product is the response, the batch is the treatment, and the day is the block. Using our procedure, we believe that we could diminish the block effect of days in the analysis and come up with an appropriate pairwise multiple comparison between the batches.

@@

Based on this, we reject our original hypothesis that using any accelerant is better than no accelerant in favor of the theory that 75.5% ethanol alcohol has a negative effect on burning distance.

>>>>

---
title: "Accelerating Burn Speed of Birch Wood Popsicle Sticks"
author: "E. Croffoot-Suede, S. Kim"
date: "May 2, 2022"
output:
  pdf_document:
    pandoc_args: --listings
    includes:
      before_body: title.sty
      in_header: preamble.tex
---

@@

Curiously, all three solutions had warnings on the bottles warning of the flammability of each liquid, which boded well for our hypothesis that soaking objects in them would increase that object's flammability.

### Hypothesis

If we soak popsicle sticks in ethanol alcohol, isopropyl alcohol, and acetone and burn them, comparing distance burned against a dry popsicle stick, then we think that using any liquid accelerant would be faster than no liquid accelerant. We also think that in comparing individual treatments, there would be no significant difference between the accelerants, but there would be a significant difference between no accelerant and with accelerant.

## Study Design and Procedure

@@

For the experiment, we used birch wood popsicle sticks. We also prepared our treatments, which were the ethanol alcohol, isopropyl alcohol, acetone, and no accelerant. To hold the burning sticks off the ground (so they could be lit from below, since flames rise), we used four mason jars, one for each stick within a block. To light the sticks, we used a barbeque lighter, and for timing, we used an Apple iPhone. For safety purposes, we also had extra water.

@@

To ensure that we could find statistically significant results, we ran a power calculation. We assumed that we would use six time blocks, and we chose an alpha level for this of $\alpha = 0.05$, which is the probability of a Type I error (false rejection). We chose a Type II error probability of 0.1. Given a $k = 4$ (four treatments), the calculated sample size was 20, but given a $k = 6$ (six time blocks), our calculated sample size was 24. Given our two-way layout, we should take the larger $N$ calculated out of $k = 4, 6$. Based on this, we used $N = 24$, meaning that we should use 6 blocks of 4 observations each.

@@

This method of block generation, however, means that it is possible to have multiple iterations of one treatment and none of other treatments in the same block. For example, it is possible to have "acetone", "acetone", "isopropyl", and "dry" in a single block, meaning that "ethanol" would not be tested in this block. This gives us arbitrary, unbalanced block data.

### Statistical Tools

Data were cleaned using Python version 3.7 (Python, Wilmington, DE) and Pandas version 1.0.3. Data were analyzed using R version 4.0.3 (The R Project for Statistical Computing, Indianapolis, IN). The Skillings-Mack procedure was run in R using the NSM3 package version 1.16. The Wilcoxon rank-sum tests were run using the stats package in R.

## Statistical Analysis

@@

Figure 1 shows the range of response values, but to get a better view of the treatment effects, we also made Figure 2, a plot of unburnt distance by accelerant. We are not using the blocking factor as a factor of interest, so the only plot to be made using a factor against response is with treatment, as shown in Figure 2. We also see that in Figure 2, the variances between the groups do not appear to be the same, since acetone appears to have a much larger spread. Additionally, ethanol alcohol appears to have the highest median, whilst acetone, isopropyl alcohol, and no accelerant appear to have similar medians.

@@

Because of its higher relative asymptotic efficiency, we began by building an analysis of variance model on the data. We believe that the independence condition is met because each popsicle stick should not affect another popsicle stick during the burn. However, when we checked the conditions for an analysis of variance, we found that equal variance condition and the normality condition were not met for a parametric test.

```{R normality, fig.show="hold", out.width="50%", fig.height = 5, echo = FALSE}
plot(aov(distance ~ type + run, data = fire), which = 1, main = "Figure 3: Residuals vs Fitted")
plot(aov(distance ~ type + run, data = fire), which = 2, main = "Figure 4: Normal Q-Q")
```

@@

In Figure 4, there are flaired tails that indicate an overdispersed data distribution compared to a normal distribution. That is, the distribution was "squished" toward the center compared to the normal distribution.

@@

Thus, our model is updated to

@@

The Skillings-Mack procedure requires that the $N$ random variables are mutually independent. That is, $X_{ij1}, ..., X_{ijc_{ij}}$, for $i = 1, ..., n$, and $j = 1, ..., k$ are mutually independent. We believe this to be true for our data given the collection method. It also requires that for each fixed $(i, j)$, the $c_{ij}$ random variables are a random sample from a continuous distribution function $F_{ij}$. We also believe this to be true because we used a random sample out of the population of popsicle sticks in the world. Lastly, the treatment response should be a continuous variable such that "less than", "equal to", and "greater than" can be placed into context. We believe that "length" in millimeters as a response variable satisfies this condition.

@@

When we ran the Skillings-Mack procedure, we found an asymptotic upper-tail probability of $0.0536 < 0.1 = \alpha$, and a Skillings-Mack SM test statistic of 7.6587 on 3 degrees of freedom. Given this, we concluded that there is a difference in the medians between at least one of the groups.

In order to find which groups have a significant difference, we need to use an associated pairwise comparison procedure that can also be used against arbitrary incomplete block designs. To determine this, we created a novel pairwise multiple comparisons procedure that can be used on arbitrary incomplete block designs:

1. Run a Wilcoxon rank-sum test on the data for all $6 = {4\choose2}$ treatment pairwise comparisons, where $H_0: F(t) = G(t)$ and $H_A: F(t) \neq G(t)$, where $F(t)$ represents the cumulative distribution function for one treatment, and $G(t)$ represents the cumulative distribution function for another treatment
2. Keep the block and response values together, but randomly assign a new treatment for each observation, then calculate the Wilcoxon rank sum for all $6 = {4\choose2}$ treatment pairwise comparisons. This will spread the block effect across all treatments. Repeat this $r$ times, and store each of the test statistics for the pairwise comparisons for each of these tests.
3. For each of the six pairwise comparisons groups, compare the Bonferroni-adjusted \nth{5} and \nth{95} percentiles of the $r$ simulated random-treatment Wilcoxon rank-sum test statistics against the test statistic from each of the pairwise Wilcoxon tests from the actual data. If the actual Wilcoxon rank-sum test statistic falls inside of the Bonferroni-adjusted (0.05, 0.95) percentile from the random-treatment distribution, there is a significant difference between the two groups.

This procedure is used to determine, given two groups (in this case, each pair of treatments), if there is strong statistical evidence to support the hypothesis that one of the groups is significantly different from the other. Because we are using a nonparametric method, we are comparing the medians. This diminishes the effect of using non-normal data. The Wilcoxon rank-sum test calculates distance from the medians using the sum of the rank, not the actual value's difference from the median, which also helps to diminish the effect of extraordinarily large or small outliers.

@@

We see that the spread for acetone reaches the lowest distance and has the lowest minimum. Because of this, to minimize unburnt distance or to maximize burnt distance, we would want to use acetone or any accelerant not statistically different from acetone.

@@

However, we believe that the novel pairwise multiple comparisons procedure that we designed would still be valid given data obtained in an arbitrary incomplete block design. For example, suppose we were looking at data for a factory that produces products in batches, but which does not produce every batch every day. This would generate data in an arbitrary incomplete block design, where a metric of the product is the response, the batch is the treatment, and the day is the block. Using our procedure, we believe that we could diminish the block effect of days in the analysis and come up with an appropriate pairwise multiple comparison between the batches.

@@

Based on this, we reject our original hypothesis that using any accelerant is better than no accelerant in favor of the theory that acetone and isopropyl alcohol burn just as well as no accelerant.
